---
title: "Project Overview"
author: "tobiasst"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    df_print: paged
    code_folding: hide
---

```{r init, message=F, warning=F}
library(SummarizedExperiment)
library(biomaRt)
rerun <- T
```

## build sample table from runinfo table

```{r samples}
samples <- read.delim("data/cluster/SraRunTable.txt")
```

Li R, Grimm SA, Mav D, Gu H et al. Transcriptome and DNA Methylome Analysis in a Mouse Model of Diet-Induced Obesity Predicts Increased Risk of Colorectal Cancer. Cell Rep 2018 Jan 16;22(3):624-637. PMID: 29346762
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE85731


Data is paired-end, stranded. RNA-Seq was performed in colonic epithelium from mice on three different dietary regimens as follows: 1) low-fat diet for 20 weeks; 2) high-fat diet for 20 weeks; 3) high-fat diet for an initial 15 weeks then switched to low-fat diet for another 5 weeks.

## Collect data from cluster 

```{r collect_data}

if (rerun) {
  cluster.dir <- "/Users/tobiasst/Desktop/mount/work/project/tobias/wade_2018/"
  
  cmat <- c()
  tmat <- c()
  
  for (sample in samples$Run) {
    sample.name <- 
      starcounts <- read.delim(paste(cluster.dir,sample,"_out/",sample,".ReadsPerGene.out.tab",sep=""), header=F)  
    cmat <- cbind(cmat, starcounts[,4])
    if (!sum(starcounts[,2])==0) {
      rsemtpm <- read.delim(paste(cluster.dir,sample,"_out/",sample,"_rsem.genes.results",sep=""), header=T)
      tmat <- cbind(tmat, rsemtpm$TPM)
    } else {
      tmat <- cbind(tmat, rep(NA, nrow(tmat)))
    }
    
  }
  rownames(cmat) <- starcounts[,1]
  rownames(tmat) <- rsemtpm$gene_id
  
  flt <- !(grepl("^N_", rownames(cmat)))
  cmat <- cmat[flt,]
  
  colnames(cmat) <- samples$Run
  colnames(tmat) <- samples$Run
  tmat <- tmat[match(rownames(cmat), rownames(tmat)),]
  
  dim(cmat)
  flt <- rowSums(cmat) > 0
  cmat <- cmat[flt,]
  tmat <- tmat[flt,]
  dim(cmat)
  
  samplesM <- data.frame(group=samples$diet)
  rownames(samplesM) <- samples$Run
  samplesM$diet[samplesM$group=="10% fat diet (LF)"] <- "LF"
  samplesM$diet[samplesM$group=="60% fat diet (HF)"] <- "HF"
  samplesM$diet[samplesM$group=="HF diet then switched to LF diet"] <- "HF_LF"
  samplesM$diet <- factor(samplesM$diet)
  
  ## get Annotation data
  ensembl <- useDataset("mmusculus_gene_ensembl", useMart("ensembl"))
  bm <- getBM(filters = "ensembl_gene_id",
              attributes = c("ensembl_gene_id","entrezgene","external_gene_name","gene_biotype","description"),
              values = rownames(cmat),
              mart = ensembl)
  
  annot <- data.frame(id=rownames(cmat))
  annot$symbol <- bm$external_gene_name[match(rownames(cmat), bm$ensembl_gene_id)]
  annot$entrez <- bm$entrezgene[match(rownames(cmat), bm$ensembl_gene_id)]
  annot$biotype <- bm$gene_biotype[match(rownames(cmat), bm$ensembl_gene_id)]
  annot$description <- bm$description[match(rownames(cmat), bm$ensembl_gene_id)]
  
  se <- SummarizedExperiment(assays=list(counts=cmat, tpms=tmat), colData=samplesM, rowData=annot)
  dim(se)
  
  saveRDS(se, file="se.rds")
} else {
  se <- readRDS("se.rds")
}

```

## Read distribution

```{r fig_counts, fig.width=5, fig.height=5}
par(mar=c(6,4,2,2)+0.1)
total.counts <- apply(assays(se)$counts,2,sum)
o <- order(total.counts)
total.counts <- total.counts[o]
barplot(total.counts, las=2, col=se$group[o])
```

## Genes detected

```{r fig_genes, fig.width=5, fig.height=5}
par(mar=c(6,4,2,2)+0.1)
total.genes <- apply(assays(se)$counts,2,function(x){sum(x>0)})
o <- order(total.genes)
total.counts <- total.genes[o]
barplot(total.counts, las=2, col=se$group[o])

```

## PCA

```{r pca, fig.width=6, fig.height=6}
mat <- log2(assays(se)$tpms+0.01)
mat <- mat[apply(mat,1,var)>0,]
pca <- prcomp(t(mat))
df <- data.frame(pca$x[,1:4])
plot(df, col=se$group)
```

```{r}
acan <- "ENSMUSG00000030607"
assays(se)$tpms[acan,]
assays(se)$tpms["ENSMUSG00000000001",]

plot(log(assays(se)$tpms[,1]+0.01), log(assays(se)$counts[,1]+1))
```

```{r}
sessionInfo()
```

